Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-06-19 src/tex/ms.tex
Update

Many data analysis tasks in today's gravitational wave (GW) astrophysics, such
as match filtering search and parameter estimation, rely on accurate waveform
models. Because using numerical relativity (NR) waveforms in these tasks is
prohibitively expensive, the community has constructed waveform approximants
that can be evaluated much faster. There are three families of commonly used GW
approximants, which are effective-one-body (EOB) \citep{ossokine2020multipolar,
cotesta2020frequency, taracchini2014effective}, Numerical Relativity (NR)
surrogate \citep{islam2022surrogate, varma2019surrogate, varma2019surrogate2}
and phenomenological (Phenom) models \citep{husa2016frequency,
khan2016frequency, garcia2020multimode, pratten2021computationally}. While the
detail of construction of these models are different, they all have a set of
internal parameters that can be calibrated to NR waveforms. The quality of a
waveform model is determined by the ansatz used and the accuracy of the
calibrated parameters.

@@

The LIGO-VIRGO-KAGRA (LVK) collaboration
(\citep{LIGOScientific:2014pky,LIGOScientific:2021usb,LIGOScientific:2021djp,
VIRGO:2014yos,KAGRA:2020tym}.) has recently started their fourth observational
run on May 26, 2023. This new run is expected to double the rate of current binary black
hole (BBH) observations \citep{abbott2020prospects}. Additionally, the
sensitivity of interferometers will be increased to capture more details of GW.
Having instruments with higher sensitivity, GW models of equal or higher
accuracy than observations should be used to extract GW information. Otherwise,
the extracted information would be affected more by GW models instead of
interferometer sensitivity, resulting in a bottleneck in GW analyses. Although
GW models are accurate enough for current analyses, the accuracy of current
models will no longer suffice for future data analyses
\citep{purrer2020gravitational}. Hence, it is necessary for us to develop and
improve GW models.

@@

Automatic differentiation (AD) is a method to calculate derivatives of functions
up to machine precision. In traditional numerical calculations, derivatives are
usually obtained through numerical derivatives. Symbolic derivatives were
available but it was less efficient. Both methods were not viable in machine
learning, where back-propagation requires precise and rapid derivative
calculations. In {\python}, packages including \texttt{pytorch} \citep{pytorch},
\texttt{tensorflow} \citep{tensorflow2015-whitepaper}, etc. utilizes AD to train
machine learning models. AD's algorithm is intuitive in nature. Functions
defined are decomposed into tree structures of primitives, such as addition or
function evaluations. Since these operations are fundamental, they were saved as
pairs internally. Differentiation proceeds forward following the tree structure,
with the application of the chain rule in each step to evaluate its derivative.
Analytic derivatives of such operations are applied in each step and the desired
derivative can then be obtained by composing back the original function
according to the original structure. {\ripple} \citep{ripple} was a new
implementation of IMRPhenomD, one of the models within the Phenom family. It was first implemented
in \texttt{lalsuite} using \texttt{C}. In order to make use of AD, it was
rewritten using \jax, a {\python} package that supports AD. Using
{\ripple}, one can apply AD to GW models to obtain precise derivatives, thus
allowing one to freely use derivative-based algorithms to perform data analyses. 
% Need to cite references for AD

% Put more focus on the joint optimization instead of having a set of more accurate 
% coefficients, since the method is the main point
In this paper, we investigate the possibility of further improving the accuracy
of IMRPhenomD by jointly optimizing all the fitting coefficients given NR
waveforms, and what constraints one may face when trying to further improve. We
find that simply by applying gradient descent algorithm, one can obtain a better
set of waveform coefficients, thus improving the accuracy of the model.
Furthermore, by comparing the accuracy of optimized and original waveforms, we
find that model-generated waveforms are very sensitive to their source
parameters. Specifically, IMRPhenomD favors certain parts of the parameter
space. This means IMRPhenomD introduces systematic bias to other GW analysis
tasks. This showcases the flaws of the ansatz and allows us to have a deeper
understanding of Phenom models.

@@

>>>>

Many data analysis tasks in today's gravitational wave (GW) astrophysics, such
as match filtering search \citep{fill} and parameter estimation\citep{fill}, rely on accurate waveform
models. Because using numerical relativity (NR) waveforms in these tasks is
prohibitively expensive, the community has constructed waveform approximants
that can be evaluated much faster. There are three families of commonly used GW
approximants, which are effective-one-body (EOB) \citep{ossokine2020multipolar,
cotesta2020frequency, taracchini2014effective}, Numerical Relativity (NR)
surrogate \citep{islam2022surrogate, varma2019surrogate, varma2019surrogate2}
and phenomenological (Phenom) models \citep{husa2016frequency,
khan2016frequency, garcia2020multimode, pratten2021computationally}. While the
detail of construction of these models are different, they all have a set of
internal parameters that can be calibrated to NR waveforms. The quality of a
waveform model is determined by the ansatz used and the accuracy of the
calibrated parameters.

@@

The LIGO-VIRGO-KAGRA (LVK) collaboration
(\citep{LIGOScientific:2014pky,LIGOScientific:2021usb,LIGOScientific:2021djp,
VIRGO:2014yos,KAGRA:2020tym}.) has recently started their fourth observational
run on May 26, 2023. Because of the sensitivity improvement, the new run is
expected to double the size of current binary black hole (BBH) observations
\citep{abbott2020prospects}. The improved sensitivity also implies we expect to
detect individual events with a higher signal-to-noise ratio (SNR). This means
we can resolve more features in the signal, at the same time put a more
stringent requirement on the accuracy of our waveform model\citep{fill}.

@@

Due to the number of calibration parameters (often in a couple hundreds if not
more), waveform models is usually calibrated in pieces \citep{fill}. This
ignores the correlation between different parts of the waveform model and limits
it quality. Recently, there has been efforts in rebuilding waveform models
\citep{fill} that supports automatic differentiation (AD)\citep{fill}, which is
a technique used to compute derivatives of functions down to machine precision
without issues of scaling up to high dimension or expression swelling. In
particular, {\ripple} \citep{ripple} exposes the calibration parameters to the
user. This allows us to make use of infrastructures that are heavily used in
machine learning, such as gradient descent and back propagation\cite{fill}, to
improve the calibration of the waveform models.

% Put more focus on the joint optimization instead of having a set of more accurate 
% coefficients, since the method is the main point
In this paper, we investigate the possibility of further improving the accuracy
of a waveform model, IMRPhenomD\citep{fill}, by jointly optimizing all the
calibration coefficients given a set of NR waveforms. Training on a small number
of NR waveforms, we demonstrate one can improves the match between IMRPhenomD
and NR waveforms over a decently sized parameter space, up to mass ratio $q=8$
We also explore how different parts of the source parameter space (e.g. the
primary and second spins) respond to the optimization procedure, by optimizing
the waveform separately for different regions of the parameter space. This gives
us hints on whether the waveform model performs equally well in different
regions of the parameter space.

@@

% Moved down from intro
% In particular, we find it is easier IMRPhenomD
% favors certain parts of the parameter space. This means IMRPhenomD introduces
% systematic bias to other GW analysis tasks. This showcases the flaws of the
% ansatz and allows us to have a deeper understanding of Phenom models.
